Action Groupoids and Fractional Types

### 2016-06-10 pibackground.tex
done with sec 3 for now; will start sec 4

The main syntactic vehicle for the technical developments in this
paper is the language $\Pi$ whose only computations are isomorphisms
between finite types and equivalences between these
isomorphisms~\cite{James:2012:IE:2103656.2103667,Carette2016}. We
present the syntax and operational and denotational semantics of
the parts of the language relevant to our development.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

We give an operational semantics for $\Pi$.  Operationally, the
semantics consists of a pair of mutually recursive evaluators that
take a combinator and a value and propagate the value in the forward
direction~$\triangleright$ or in the backward
direction~$\triangleleft$. We show the complete forward evaluator in
Fig.~\ref{opsem}; the backward evaluator is easy to infer.

As an example, let $\mathbb{3}$ abbreviate the type
$(\ot \oplus \ot) \oplus \ot$ and consider the following six
combinators of type $\mathbb{3} \iso \mathbb{3}$:

\[\begin{array}{rcl}
a_1 &=& \idiso \\
a_2 &=& \swapp \oplus \idiso \\
a_3 &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
a_4 &=& a_2 \odot a_3 \\
a_5 &=& a_3 \odot a_2 \\
a_6 &=& a_4 \odot a_2
\end{array}\]

There are three values of type $\mathbb{3}$ which are
$\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
$\inr{\unitv}$. Evaluting $a_2$ on each of these values, we get:
\[\begin{array}{rcl}
\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
&=& \inl{\inr{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
&=& \inl{\inl{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
&=& \inr{\unitv}
\end{array}\]

Every 1-combinator has an \emph{order} $n$: applying this combinator
$n$ times is equivalent to the identityA couple of examples: compute order
The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
6 automorphisms on it:

The combinators realize all the 3! permutations on the set of
cardinality 3. The permutation $a_1$ is the identity permutation; the
permutations $a_2$, $a_3$, and $a_6$ swap two of the three elements
leaving the third intact; the permutations $a_4$ and $a_5$ rotate the
three elements. A crucial property of combinators is their
\emph{order} which classifies them according to the number of
iterations needed for them to compose to the identity. In our case,
the reader is invited to check the following: $\mathit{order}(a_1)=1$,
$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
$\mathit{order}(a_4)=\mathit{order}(a_5)=3$. Generally speaking to
calculate the order of a combinator $c$, we apply $c$ to every value
and observe the length of the orbit. The order is the lcm of all the
orbits.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Groupoid Model}

For $c_1, c_2 : \tau_1\iso\tau_2$, we have level-2 combinators $\alpha
: c_1 \isotwo c_2$ which are equivalences of isomorphisms, and which
happen to correspond to the coherence conditions for rig
groupoids. Many of the level 2 combinators not used. We only present
the ones we use in this paper.

 From the perspective of category theory, the language $\Pi$ models what is
called a \emph{symmetric bimonoidal category} or a \emph{commutative rig
category}. These are categories with two binary operations and satisfying the
axioms of a commutative rig (i.e., a commutative ring without negative
elements also known as a commutative semiring) up to coherent
isomorphisms. And indeed the types of the $\Pi$-combinators are precisely the
commutative semiring axioms. A formal way of saying this is that $\Pi$ is the
\emph{categorification}~\cite{math/9802029} of the natural numbers. A simple
(slightly degenerate) example of such categories is the category of finite
sets and permutations in which we interpret every $\Pi$-type as a finite set,
interpret the values as elements in these finite sets, and interpret the
combinators as permutations.

Explain that we only reach a trivial class of groupoids

introduce discreteC and discreteG

discrete category: give a set $S$ objects are $S$, morphisms are only
identity morphisms; each object has exactly one morphism id. That is a
groupoid, trivially, because id is symmetric

Instead of modeling the types of $\Pi$ using sets and the combinators using
permutations we use a semantics that identifies $\Pi$-combinators with
\emph{paths}. More precisely, we model the universe of $\Pi$-types as a space
\AgdaFunction{U} whose points are the individual $\Pi$-types (which are
themselves spaces \AgdaBound{t} containing points). We then postulate that
there is path between the spaces \AgdaBound{t₁} and \AgdaBound{t₂} if there
is a $\Pi$-combinator $c : t_1 \iso t_2$. Our postulate is similar in spirit
to the univalence axiom but, unlike the latter, it has a simple computational
interpretation. A path directly corresponds to a type isomorphism with a
clear operational semantics as presented in the previous section. As we will
explain in more detail below, this approach replaces the datatype
\AgdaSymbol{≡} modeling propositional equality with the datatype
\AgdaSymbol{⟷} modeling type isomorphisms. With this switch, the
$\Pi$-combinators of Fig.~\ref{pi-combinators} become \emph{syntax} for the
paths in the space $U$. Put differently, instead of having exactly one
constructor \AgdaInductiveConstructor{refl} for paths with all other paths
discovered by proofs (see Secs. 2.5--2.12 of the HoTT
book~\citeyearpar{hottbook}) or postulated by the univalence axiom, we have
an \emph{inductive definition} that completely specifies all the paths in the
space $U$.

We begin with the datatype definition of the universe \AgdaFunction{U} of finite
types which are constructed using \AgdaInductiveConstructor{ZERO},
\AgdaInductiveConstructor{ONE}, \AgdaInductiveConstructor{PLUS}, and
\AgdaInductiveConstructor{TIMES}. Each of these finite types will correspond
to a set of points with paths connecting some of the points. The underlying
set of points is computed by \AgdaSymbol{⟦}\_\AgdaSymbol{⟧} as follows:
\AgdaInductiveConstructor{ZERO} maps to the empty set~\AgdaSymbol{⊥},
\AgdaInductiveConstructor{ONE} maps to the singleton set \AgdaSymbol{⊤},
\AgdaInductiveConstructor{PLUS} maps to the disjoint union \AgdaSymbol{⊎},
and \AgdaInductiveConstructor{TIMES} maps to the cartesian product
\AgdaSymbol{×}.

We note that the refinement of the $\Pi$-combinators to combinators on
pointed spaces is given by an inductive family for
\emph{heterogeneous} equality that generalizes the usual inductive
family for propositional equality. Put differently, what used to be
the only constructor for paths \AgdaInductiveConstructor{refl} is now
just one of the many constructors (named
\AgdaInductiveConstructor{id⟷} in the figure). Among the new
constructors we have \AgdaInductiveConstructor{◎} that constructs path
compositions. By construction, every combinator has an inverse
calculated as shown in Fig.~\ref{sym}. These constructions are
sufficient to guarantee that the universe~\AgdaFunction{U} is a
groupoid. Additionally, we have paths that connect values in different
but isomorphic spaces like \ldots

The example \AgdaFunction{notpath} which earlier required the use of the
univalence axiom can now be directly defined using
\AgdaInductiveConstructor{swap1₊} and \AgdaInductiveConstructor{swap2₊}. To
see this, note that \AgdaPrimitiveType{Bool} can be viewed as a shorthand for
\AgdaInductiveConstructor{PLUS} \AgdaInductiveConstructor{ONE}
\AgdaInductiveConstructor{ONE} with \AgdaInductiveConstructor{true} and
\AgdaInductiveConstructor{false} as shorthands for
\AgdaInductiveConstructor{inj₁} \AgdaInductiveConstructor{tt} and
\AgdaInductiveConstructor{inj₂} \AgdaInductiveConstructor{tt}. With this in
mind, the path corresponding to boolean negation consists of two ``fibers'',
one for each boolean value as shown below:

\noindent In other words, a path between spaces is really a collection of
paths connecting the various points. Note however that we never need to
``collect'' these paths using a universal quantification.

Shouldn't we also show that \AgdaPrimitiveType{Bool} contains exactly
$2$ things, and that TRUE and FALSE are ``different''? The other thing
is, whereas not used to be a path between Bool and Bool, we no longer
have that.  Shouldn't we show that, somehow, BOOL and BOOL.F 'union'
BOOL.T are somehow ``equivalent''?  And there there is a coherent
notpath built the same way?  Especially since I am sure it is quite
easy to build incoherent sets of paths!

>>>>

The main syntactic vehicle for the technical developments in this
paper is the language $\Pi$ whose only computations are isomorphisms
between finite types and equivalences between these
isomorphisms~\cite{James:2012:IE:2103656.2103667,Carette2016}. We
present the syntax and operational semantics of the parts of the
language relevant to our development.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics and the Order of 1-Combinators}

We give an operational semantics for the 1-combinators of $\Pi$ which
represent ``conventional'' programs.  Operationally, the semantics
consists of a pair of mutually recursive evaluators that take a
combinator and a value and propagate the value in the forward
direction~$\triangleright$ or in the backward
direction~$\triangleleft$. We show the complete forward evaluator in
Fig.~\ref{opsem}; the backward evaluator is easy to infer.

As an example, let $\mathbb{3}$ abbreviate the type
$(\ot \oplus \ot) \oplus \ot$. There are three values of type
$\mathbb{3}$ which are $\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
$\inr{\unitv}$. There are six combinators of type
$\mathbb{3} \iso \mathbb{3}$, each representing a different
permutation of three elements, which could be written as follows:
\[\begin{array}{rcl}
a_1 &=& \idiso \\
a_2 &=& \swapp \oplus \idiso \\
a_3 &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
a_4 &=& a_2 \odot a_3 \\
a_5 &=& a_3 \odot a_2 \\
a_6 &=& a_4 \odot a_2
\end{array}\]
Tracing the evaluation of $a_2$ on each of the possible inputs yields:
\[\begin{array}{rcl}
\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
&=& \inl{\inr{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
&=& \inl{\inl{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
&=& \inr{\unitv}
\end{array}\]
Thus the effect of combinator $a_2$ is to swap the values
$\inl{\inl{\unitv}}$ and $\inl{\inr{\unitv}}$ leaving the value
$\inr{\unitv}$ intact. Iterating $a_2$ again is therefore equivalent
to the identity permutation, which can be verified using the full set of
2-combinators~\cite{Carette2016}:
\[\begin{array}{rcl}
a_2 \odot a_2 &=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\
&\isotwo& (\swapp \odot \swapp) \oplus (\idiso \odot \idiso) \\
&\isotwo& \idiso \oplus \idiso \\
&=& \idiso
\end{array}\]

A similar calculation shows that the combinator $a_1$ is the identity
permutation; the combinators $a_3$ and $a_6$ also swap two of the
three elements leaving the third intact; and the combinators $a_4$ and
$a_5$ rotate the three elements. We therefore have
$\mathit{order}(a_1)=1$,
$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.
